Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-11 Paper/application.tex
polish application, conclusion, and related work

  We can extend the above construction to model both singleton and replace-all operations. In this paper, we only illustrate the construction of replace-all in Figure \ref{fig:replace-all}. The SFT $\mathcal{T}_2$ combines pattern matching and replacement using the construction just described. The SFT $\mathcal{T}_1$ matches strings that do not contain any substrings matching the pattern. Two $\varepsilon$-transitions connect these components: one from $\mathcal{T}_1$'s accepting state to $\mathcal{T}_2$'s initial state, and another from $\mathcal{T}_2$'s accepting state back to $\mathcal{T}_1$'s initial state. This cyclic structure enables the replace-all operation by repeatedly applying pattern replacement until no more matches exist in the input string.

@@

We evaluate {CertiStrR} using benchmarks from SMT-LIB 2025~\cite{smtlib_benchmarks}, focusing on the \textsf{QF\_S} and \textsf{QF\_SLIA} logic fragments. The benchmarks are divided into two categories: (1) singleton matching and (2) replace-all matching. In addition to replacement, they also include string concatenation and regular constraints in the tests.

@@

\begin{table*}[h]
\centering
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{llccccccc}
  \toprule
  \textbf{Subgroup} & \textbf{Test} & \textbf{Count} & \textbf{SAT} & \textbf{UNSAT} & \textbf{Inconclusive} & \textbf{CertiStrR (s)} & \textbf{Ostrich (s/timeout)} & \textbf{CVC5 (s/timeout)} \\
  \midrule
  \multirow{2}{*}{Replace Singleton} & Replace RE  & 94  & 77  & 16  & 1 & 0.79 & 166.27/27 & 2.67/20 \\
   & Replace Str & 323 & 103 & 211 & 9 & 1.94 & 7.10/20 & 0.81/1 \\
  \multirow{4}{*}{Replace All} & PCP & 1000 & 751 & 249 & 0 & 0.14 & 3.412/0 & unknown \\
   & RNA SAT   & 500  & 500 & 0   & 0 & 2.81 & 55.61/0 & unknown \\
   & RNA UNSAT & 500  & 0   & 500 & 0 & 3.14 & 64.185/0 & unknown \\
   & WebApp      & 255  & 40  & 215 & 0 & 0.79 & 23.67/17 & 8.379/1 \\
  \bottomrule
\end{tabular}
\caption{Experimental results}
\label{tab:string_operations}
\end{table*}

@@

\paragraph{Transducers vs. Replacement Operations.}  
Although both transducers and replacement operations can express string transformations for security-related verification, transducers are strictly more expressive because they can capture dependencies between the input pattern and the replacement string. In contrast, all replacement operations in SMT-LIB treat the pattern and the replacement string as independent. For example, in the SFT shown in Figure~\ref{fig:html-sanitizer-sft}, a transition may be defined over an interval such as $['a'-'z']$, and the output function computes its output directly based on the specific input character. Replacement operations, however, cannot express such dependencies.

Consider the transition $['a'-'z']/id$, where $id$ is the identity function. To model this transition with replacement operations, the interval must be explicitly expanded into the individual characters $\texttt{'a'}$ through $\texttt{'z'}$, requiring 26 nested \texttt{replace-all} operations—one for each character. Thus, modeling $['a'-'z']/id$ with replacement operations yields an expression of the form:
\[
\texttt{replace}(\texttt{replace}(\ldots,~\texttt{"b"},~\texttt{"b"},~g),~\texttt{"a"},~\texttt{"a"},~g).
\]

When the interval contains a large set of characters—as is common in web applications that support the full Unicode range of $0x10FFFF$ code points—this nested structure becomes impractical, leading to significant inefficiency and complexity in the replacement-based approach.

@@

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{lccc}
      \toprule
      & \textbf{Defs} & \textbf{Lemmas} & \textbf{Proofs (LOC)} \\
      \midrule
      Abstract & 17 & 21 & 3274 \\
      Implementation & 51 & 43 & 2700 \\
      Interval & 15 & 29 & 1500 \\
      \bottomrule
  \end{tabular}
  \caption{Overview of the effort of certified development}
  \label{tab:abstract_impl}
\end{table}

>>>>

  We can extend the above construction to model both replace-first and replace-all operations. In this paper, we only illustrate the construction of replace-all in Figure \ref{fig:replace-all}. The SFT $\mathcal{T}_2$ combines pattern matching and replacement using the construction just described. The SFT $\mathcal{T}_1$ matches strings that do not contain any substrings matching the pattern. Two $\varepsilon$-transitions connect these components: one from $\mathcal{T}_1$'s accepting state to $\mathcal{T}_2$'s initial state, and another from $\mathcal{T}_2$'s accepting state back to $\mathcal{T}_1$'s initial state. This cyclic structure enables the replace-all operation by repeatedly applying pattern replacement until no more matches exist in the input string.

@@

We evaluate {CertiStrR} using benchmarks from SMT-LIB 2025~\cite{smtlib_benchmarks}, focusing on the \textsf{QF\_S} and \textsf{QF\_SLIA} logic fragments. The benchmarks are divided into two categories: (1) replace-first matching and (2) replace-all matching. In addition to replacement, they also include string concatenation and regular constraints in the tests.

@@

\begin{table*}[h]
\centering
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{llccccccc}
  \toprule
  \textbf{Subgroup} & \textbf{Test} & \textbf{Count} & \textbf{SAT} & \textbf{UNSAT} & \textbf{Inconclusive} & \textbf{CertiStrR (s)} & \textbf{Ostrich (s/timeout)} & \textbf{CVC5 (s/timeout)} \\
  \midrule
  \multirow{2}{*}{Replace-First} & Replace RE  & 94  & 77  & 16  & 1 & 0.79 & 166.27/27 & 2.67/20 \\
   & Replace Str & 323 & 103 & 211 & 9 & 1.94 & 7.10/20 & 0.81/1 \\
  \multirow{4}{*}{Replace-All} & PCP & 1000 & 751 & 249 & 0 & 0.14 & 3.412/0 & unknown \\
   & RNA SAT   & 500  & 500 & 0   & 0 & 2.81 & 55.61/0 & unknown \\
   & RNA UNSAT & 500  & 0   & 500 & 0 & 3.14 & 64.185/0 & unknown \\
   & WebApp      & 255  & 40  & 215 & 0 & 0.79 & 23.67/17 & 8.379/1 \\
  \bottomrule
\end{tabular}
\caption{Experimental results}
\label{tab:string_operations}
\end{table*}

@@

\paragraph{Transducers vs. Replacement Operations.}  
Although both transducers and replacement operations can express string transformations for security-related verification, transducers are strictly more expressive because they can capture dependencies between the input pattern and the replacement string. In contrast, all replacement operations in SMT-LIB treat the pattern and the replacement string as independent. For example, in the SFT shown in Figure~\ref{fig:html-sanitizer-sft}, a transition may be defined over an interval such as $[\texttt{a-z}]$, and the output function computes its output directly based on the input character. Replacement operations, however, cannot express such dependencies.
Because the replacement string is not a function.

Consider the transition $[\texttt{a-z}]/id$, where $id$ is the identity function. To model this transition with replacement operations, the interval must be explicitly expanded into the individual characters $\texttt{'a'}$ through $\texttt{'z'}$, requiring 26 nested \texttt{replace-all} operations—one for each character. Thus, modeling $[\texttt{a-z}]/id$ with replacement operations yields an expression of the form:
\[
\texttt{replace}(\texttt{replace}(\ldots,~\texttt{"b"},~\texttt{"b"},~g),~\texttt{"a"},~\texttt{"a"},~g).
\]

When the interval contains a large set of characters—as is common in web applications that support the full Unicode range of 
\texttt{0x10FFFF} code points—this nested structure becomes impractical, leading to significant inefficiency and complexity in the replacement-based approach.

@@

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{lccc}
      \toprule
      & \textbf{Definitions} & \textbf{Lemmas} & \textbf{Proofs (LOC)} \\
      \midrule
      Abstract & 17 & 21 & 3274 \\
      Implementation & 51 & 43 & 2700 \\
      Interval & 15 & 29 & 1500 \\
      \bottomrule
  \end{tabular}
  \caption{Overview of the effort of certified development}
  \label{tab:abstract_impl}
\end{table}

### 2025-09-11 Paper/conclusion.tex
polish application, conclusion, and related work

In this paper, we have presented the first formalization of symbolic finite transducers in the proof assistant Isabelle/HOL. Our formalization provides flexible interfaces that facilitate diverse applications through two key features: support for $\varepsilon$-transitions in both inputs and outputs, and extensibility to various boolean algebras via the refinement framework.

To demonstrate the practical utility of our formalization, we developed CertiStrR, an extension of CertiStr \cite{cpp/KanLRS22}. This implementation adds support for string replacement operations and has been evaluated on benchmarks from SMT-LIB 2024 \cite{smtlib_benchmarks}. The experimental results confirm both the efficiency and effectiveness of our approach. While we focused on string solving applications, our extensible framework for transition labels is broadly applicable to other domains, including program verification.

Future work includes extending our symbolic formalization in two key directions:
%
(1) Enriching the boolean algebra framework to support more complex theories, such as arithmetic constraints.
(2) Incorporating prioritized transitions into SFTs, which would enable precise modeling of SMT-LIB's standard semantics for replacement operations, particularly the first-match behavior.
These extensions will further enhance the expressiveness and practical applicability of our formalization while maintaining its formal verification guarantees.

>>>>

In this paper, we presented the first formalization of symbolic finite transducers (SFTs) in Isabelle/HOL. Our framework offers flexible interfaces that enable diverse applications, featuring support for $\varepsilon$-transitions on both inputs and outputs, as well as extensibility to various Boolean algebras through a refinement framework.

To demonstrate the practical utility of our formalization, we applied SFTs to model string sanitization functions commonly used in web applications and showed how to verify the correctness of these sanitizers. Furthermore, we developed CertiStrR, an extension of CertiStr \cite{cpp/KanLRS22}, which adds support for string replacement operations. We evaluated CertiStrR on benchmarks from SMT-LIB 2025 \cite{smtlib_benchmarks}, and the experimental results confirm both the efficiency and effectiveness of our approach.
